---
title: "Summarizing many univariate models"
description: |
  A secret weapon for when you're building hierarchical models.
execute:
  freeze: true
format:
  html:
    code-tools: true
editor_options: 
  chunk_output_type: console
---

We've already looked at univariate models. When we fit the same model to multiple different groups, we don't expect the same values for all the coefficients. Each thing we are studying will respond to the same variable in different ways.

Hierarchial models represent a way to *model* this variation, in ways that range from simple to complex.

Before we dive in with hierarchical structure, let's build a bridge between these two approaches.

This is useful to help us understand what a hierarchical model does.

However it is also useful from a strict model-building perspective -- so useful that [Andrew Gelman calls it a "Secret Weapon"](https://statmodeling.stat.columbia.edu/2005/03/07/the_secret_weap/)

## Loading models and data

```{r setup}
suppressPackageStartupMessages(library(dplyr))
library(ggplot2)
library(tidyr)
suppressPackageStartupMessages(library(brms))
library(tidybayes)

data(mite, package = "vegan")
data("mite.env", package = "vegan")
data("mite.xy", package = "vegan")
```

And some quick data restructuring to combine both.

```{r datasetup}
# combine data and environment
mite_data_long <- bind_cols(mite.env, mite) |> 
  pivot_longer(Brachy:Trimalc2, names_to = "spp", values_to = "abd")
```

To keep things simple and univariate, let's consider only water concentration as an independent variable.

First, a quick word about centering and scaling a predictor variable:

1.  I center the predictor by subtracting the mean. This changes the *intercept* of my linear predictor. it becomes the mean log-odds of occurrance when the water content is average
2.  I divide water content by 100. The dataset has units of **grams per Litre** of water (see `?vegan::mite.env` for more details). This is fine, but I don't think mites are able to sense differences as precise as a millimeter of water either way. by dividing by 10 I transform this into centilitres, which is more informative.

```{r smoothglm}
mite_data_long_transformed <- mite_data_long |> 
  mutate(presabs = as.numeric(abd>0),
         # center predictors
         water = (WatrCont - mean(WatrCont)) / 100
         )

mite_data_long_transformed |> 
  ggplot(aes(x = water, y = presabs)) + 
  geom_point() + 
  stat_smooth(method = "glm", method.args = list(family = "binomial")) + 
  facet_wrap(~spp)
```

some things to notice about this figure:

-   the x-axis scale has been transformed from "grams per litre" to "centilitres away from average
-   there is a ton of variation in how different species respond to water!

```{r make_mite_glms}
mite_many_glms <- mite_data_long_transformed |> 
  nest_by(spp) |> 
  mutate(logistic_regressions = list(
    glm(presabs ~ water,
        family = "binomial",
        data = data))) |> 
  mutate(coefs = list(broom::tidy(logistic_regressions)))




mite_many_glms <- mite_data_long_transformed |> 
  nest_by(spp) |> 
  mutate(logistic_regressions = list(
    glm(presabs ~ 1 + water, 
        family = binomial(link = "logit"),
        data = data)
  )) |> 
  mutate(coefs = list(broom::tidy(logistic_regressions)))

broom::tidy(mite_many_glms$logistic_regressions[[5]])
broom::glance(mite_many_glms$logistic_regressions[[5]])

mite_many_glms$coefs[[5]]



```

:::{.callout-note}
## Split-Apply-Combine

To explore this kind of thinking, we are going to use an approach sometimes called ["split-apply-combine"](https://vita.had.co.nz/papers/plyr.pdf)

There are many possible ways to do this in practice. We are using a technique here from the tidyverse, which you can [read more about](https://dplyr.tidyverse.org/articles/rowwise.html).
:::

```{r coef_plot}
mite_many_glm_coefs <- mite_many_glms |> 
  select(-data, -logistic_regressions) |> 
  unnest(coefs)

mite_many_glm_coefs |> 
  ggplot(aes(x = estimate, y = spp,
             xmin = estimate - std.error,
             xmax = estimate + std.error)) + 
  geom_pointrange() + 
  facet_wrap(~term, scales = "free")
```

As you can see, some of these estimates are high, others low. We could also plot these as histograms to see this distribution.

```{r secretweapon_hist}
mite_many_glm_coefs |> 
  ggplot(aes(x = estimate)) + 
  geom_histogram(binwidth = .5) + 
  facet_wrap(~term, scales = "free")
```

Once again, the two parameters of this model represent:

-   *Intercept* The probability (in log-odds) of a species being present at the average water concentration. some species are common, others are rare.
-   *water* this is the change in probability (in log-odds) as water increases by one centilitre per litre of substrate.

## Modelling multiple slopes in brms

The above tidyverse approach is very appealing and intuitive, but we can also do the same procedure in brms: fitting many slopes as fixed effects:

```{r, include=TRUE, eval=FALSE}
## data
glimpse(mite_data_long_transformed)



## formula
mite_water_bf <- bf(presabs ~ 0 + spp + water + water*spp,
                    family = bernoulli(link = "logit"))

# priors
get_prior(mite_water_bf, data = mite_data_long_transformed)

mite_water_prior <- c(
  prior(normal(0,1), class = "b"))
  # prior(normal(0,1), class = "Intercept")
# )


mite_water_brm <- brm(mite_water_bf,
                      data = mite_data_long_transformed,
                      prior = mite_water_prior,
                      file = "topics/correlated_effects/mite_water_brm.rds",
                      file_refit = "on_change", 
                      refresh = 0L,
                      cores = 4)

# stancode(mite_water_brm)

```

plotting coefficients

```{r}
tidybayes::get_variables(mite_water_brm)


spp_fixed_coefs <- posterior::as_draws_rvars(mite_water_brm) |> 
  tibble::enframe() |>
  unnest(value) |> 
  filter(!(name %in% c("lprior", "lp__")))

## separate slopes and intercepts
fixed_intercepts <- spp_fixed_coefs |> 
  filter(!stringr::str_detect(name, ":|water")) |> 
  mutate(spp = stringr::str_replace(name, "b_spp", "")) |> 
  select(spp, intercept = value)

## slopes
slope_first_species <- spp_fixed_coefs |> 
  filter(name == "b_water")
## get the other slopes, and add the first species to them
other_slopes <- spp_fixed_coefs |> 
  filter(stringr::str_detect(name, ":water")) |> 
  mutate(value = value + slope_first_species$value[[1]]) |> 
  mutate(spp_nob = stringr::str_replace(name, "b_spp", ""),
         spp = stringr::str_replace(spp_nob, ":water", "")) |> 
  select(spp, slope=value)

fixed_slopes <- slope_first_species |> 
  mutate(spp = "Brachy") |> 
  select(spp, slope=value) |> 
  bind_rows(other_slopes)

fixed_intercepts |> 
  ggplot(aes(xdist = intercept, y = spp)) + 
  stat_dist_pointinterval()
```


### EXERCISE
1) Add hierarchy to both the slope AND the intercept of this model. 
Do not try to model them as correlated: the `lmer` syntax would be `y ~ 1 + water + (1 | species) + (0 + water | species)`  

```{r, include=TRUE, eval=FALSE}
## data
glimpse(mite_data_long_transformed)

## formula
mite_water_hier_bf <- bf(
  presabs ~ 1 + water + (1 | spp) + (0 + water | spp),
  family = bernoulli(link = "logit"))

# priors
get_prior(mite_water_hier_bf, data = mite_data_long_transformed)

mite_water_hier_prior <- c(
  prior(normal(0,1), class = "b"),
  prior(exponential(1), class = "sd")
)


mite_water_hier_brm <- brm(mite_water_hier_bf,
                      data = mite_data_long_transformed,
                      prior = mite_water_hier_prior,
                      file = "topics/correlated_effects/mite_water_hier_brm.rds",
                      file_refit = "on_change",
                      refresh = 1000L,
                      cores = 4)

```

2) Make a plot of the slope coefficients from both models, with and without hierarchy. 
Are they the same? How are they different?

```{r}

tidybayes::get_variables(mite_water_hier_brm)

spp_hier_coefs <- mite_water_hier_brm |> 
  tidybayes::spread_rvars(b_Intercept, b_water, r_spp[spp, coef])

## intercepts
hier_intercepts <- spp_hier_coefs |> 
  filter(coef == "Intercept") |> 
  mutate(intercept = b_Intercept + r_spp) |> 
  select(spp, intercept)

hier_slopes <- spp_hier_coefs |> 
  filter(coef == "water") |> 
  mutate(slope = b_water + r_spp) |> 
  select(spp, slope)

hier_slopes |> 
  ggplot(aes(xdist = slope, y = spp)) + 
  stat_dist_pointinterval()
```

compare the slopes

```{r}

slopes_df <- bind_rows(hier_slopes |> 
                         mutate(model = "hier"),
                       fixed_slopes |> 
                         mutate(model = "fixed"))


slopes_df |> 
  # mutate(spp = forcats::fct_reorder(spp, v))
  ggplot(aes(y = spp, xdist = slope, colour = model)) + 
  stat_dist_pointinterval(position = position_dodge(width = .4))
```


Compare the intercepts

```{r}
inters_df <- bind_rows(hier_intercepts |> 
                         mutate(model = "hier"),
                       fixed_intercepts |> 
                         mutate(model = "fixed"))


inters_df |> 
  # mutate(spp = forcats::fct_reorder(spp, v))
  ggplot(aes(y = spp, xdist = intercept, colour = model)) + 
  stat_dist_pointinterval(position = position_dodge(width = .4))
```

## Comparing models

```{r}
mite_water_brm <- add_criterion(mite_water_brm, "loo")
mite_water_hier_brm <- add_criterion(mite_water_hier_brm, "loo")

loo_compare(mite_water_brm, mite_water_hier_brm)
```

